---
title: "Reproducible Research: Peer Assessment 1"
author: "Sergei Perfilev"
date: "2/20/2021"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and exploring the data

Let's load data from the .CSV file into a dataframe, then look at its structure
and see how many missing values we have in the data frame:

```{r loadData}
actData <- read.csv("data/activity.csv")
str(actData)
lapply(actData, function(x) sum(is.na(x)))
```

So, we've got some NAs (i.e. missing values) in the 'steps' column, and 
the rest of the columns are filled with data. Let's see what ranges of values 
we have in columns:

```{r ranges}
range(actData$steps, na.rm = T)

unique(actData$interval)
range(actData$interval)

unique(actData$date)        
length(unique(actData$date)); range(actData$date)
```

Now let's take a look at the distribution of the observations in time.
We have 5-minute intervals, so the dataset should consist of 12 intervals per each hour
* 24 hours per day = 288 records per date. Is that true?

```{r dates}
table(actData$date)
```


## What is mean total number of steps taken per day?

First, we calculate the total number of steps taken per day:

```{r}
stepsPerDay <- with(actData, tapply(steps, date, sum, na.rm = TRUE))
rangePerDay <- range(stepsPerDay)
```

So, the total number of steps per day varies from `r rangePerDay[1]` 
to `r rangePerDay[2]`.

Now, we calculate the mean and median of the total number of steps taken per day:

```{r}
meanPerDay <- round(mean(stepsPerDay))
medianPerDay <- median(stepsPerDay)
```

The mean is `r meanPerDay` steps per day.
The median is `r medianPerDay` steps per day.

Finally, we can make a histogram of the total number of steps taken each day:

```{r histogram1}
hist(stepsPerDay, breaks = 22, xlim = c(0, max(stepsPerDay)*1.2), 
     main = "Histogram of steps taken each day",
     xlab = "Total number of steps taken per day",
     ylim = c(0, 20))
abline(v = medianPerDay, lwd = 2, lty = 2, col = "navy")
```


## What is the average daily activity pattern?

In order to see a pattern, we will make a time series plot 
of the 5-minute interval (x-axis) and the average number of steps taken, 
averaged across all days (y-axis):

```{r patternAnalysis1}
stepsPerInterval <- with(actData, tapply(steps, interval, mean, na.rm = TRUE))
meanSteps <- data.frame(steps = stepsPerInterval, 
                        interval = as.integer(names(stepsPerInterval)))
meanSteps$interval.time <- strptime(
    format(meanSteps$interval / 100, digits = 2),
    format = "%H.%M")
rm(stepsPerInterval)

maxSteps <- round(max(meanSteps$steps))
maxIntStart <- which.max(meanSteps$steps)
maxInterval <- c(meanSteps$interval.time[maxIntStart], 
                 meanSteps$interval.time[maxIntStart + 1])
```

```{r patternPlot1}
with(meanSteps, plot(interval.time, steps, type = "l", 
     xlab = "Time (5-minute intervals)", 
     ylab = "Steps taken per interval",
     main = "Average Daily Activity Pattern"))
points(maxInterval[1], maxSteps, cex = 2, col = "red", pch = "*")
```

The maximum number of steps per one interval = 
`r maxSteps`.

The interval with maximum number of steps is 
`r paste(as.character(maxInterval, format = "%H:%M"), collapse = " - ")`.


## Imputing missing values

The presence of NAs introduces bias into our calculations and summaries. 
More precisely, our dataset contains `r sum(is.na(actData$steps))` records 
with missing values in the 'steps' column. Let's try to impute some data 
instead of those NAs.

We will apply a very simple strategy for filling in all of the missing values
in the dataset, that is for every NA we'll substitute the mean for the 
corresponding 5-minute time interval.

We create a new dataset that is equal to the original dataset but with 
the missing data filled in:

```{r fillData}
library(dplyr)
filledData <- actData %>% 
    merge(meanSteps, by = "interval") %>% 
    mutate(steps.x = ifelse(is.na(steps.x), round(steps.y), steps.x)) %>%
    mutate(steps.y = NULL) %>%
    rename(steps = steps.x) %>%
    select(c(steps, date, interval, interval.time))
```

Let's make sure that the new dataset contains no missing values:
```{r}
lapply(filledData, function(x) sum(is.na(x)))
range(filledData$steps)
```

Finally, we make a histogram of the total number of steps taken each day 
and calculate the mean and median total number of steps taken per day. 

```{r}
stepsPerDay2 <- with(filledData, tapply(steps, date, sum))
meanPerDay2 <- round(mean(stepsPerDay2))         
medianPerDay2 <- median(stepsPerDay2)            
```

```{r histogram2}
hist(stepsPerDay2, breaks = 22, xlim = c(0, max(stepsPerDay)*1.2), 
     main = "Histogram of steps taken each day (missing values replaced)",
     xlab = "Total number of steps taken per day",
     ylim = c(0, 20))
abline(v = medianPerDay2, lwd = 2, lty = 2, col = "navy")
```

The mean is `r as.integer(meanPerDay2)` steps per day.
The median is `r as.integer(medianPerDay2)` steps per day.

These values differ from the estimates from the first part of the assignment. 
The impacts of imputing missing data on the estimates of the total daily number 
of steps are:

* we don't have days with zero steps taken anymore;

* the mean and the median have increased.


## Are there differences in activity patterns between weekdays and weekends?

In order to separate activity patterns for weekdays and weekends, we'll create 
a new factor variable in the dataset with two levels – “weekday” and “weekend” 
indicating whether a given date is a weekday or weekend day.

```{r patternAnalysis2}
filledData <- 
    filledData %>% 
    mutate(wday = ifelse(
        as.POSIXlt(filledData$date)$wday %in% c(0, 6),
        "weekend", "weekday")) %>%
    mutate(wday = factor(wday, levels = c("weekend", "weekday")))

totals <- filledData %>%
    group_by(wday, interval) %>%
    summarize(avg = mean(steps))
```

Finally, we can make a panel plot containing a time series plot 
of the 5-minute interval (x-axis) and the average number of steps taken, 
averaged across all weekday days or weekend days (y-axis).

```{r patternPlot2}
library(ggplot2)
g <- ggplot(totals, aes(x = interval, y = avg)) +
    geom_line() +
    facet_grid(wday ~ .) +
    labs(x = "Time (in 5-minute intervals)", y = "Average number of steps") +
    labs(title = "Comparison of activity patterns: weekends vs weekdays")
print(g)
```
